# Overview {.unnumbered}

## Description

This workshop aims to provide R training to analysts in Liverpool City Region Combined Authority. Over four days we will cover a number of topics, which align with the current needs of the analyst team, demonstrating how R can be used for a diverse range of day-to-day analyst tasks.

## Delivery & Structure

The course will be taught across four days. Each day will run from 10AM to 3PM. Days 1 & 2 will be hosted at the [University of Liverpool](https://www.liverpool.ac.uk/geography-and-planning/contacts-and-finding-us/), and Days 3 & 4 will be hosted at LCRCA.

| Day | Activity           | Instructor         |
|-----|--------------------|--------------------|
| 1   | R Fundamentals     | Francisco Rowe     |
| 2   | Data visualisation | Patrick Ballantyne |
| 3   | APIs & Dashboards  | Patrick Ballantyne |
| 4   | Data modeling      | Francisco Rowe     |

## Before the workshop

::: callout-important
Please make sure you download and install the most recent version of R, RStudio and Quarto on the computer that you will be using during the workshop, and install the indicated R packages -- see detailed instructions below.
:::

::: callout-note
All three software packages are open and free to use.
:::

**R**

You can download R [here](https://cran.r-project.org/). Make sure you select the appropriate version for your Operating System: Windows, MacOS (Apple silicon M1/M2 or older intel Macs). For example, if you use a macOS laptop with an M1 processor, click on 'Download R for macOS' and then, click the link to download the installer file (.pkg extension for macOS) under the header 'For Apple silicon (M1/M2) Macs'. You can then open the installer and follow the instructions that you will be prompted with. For Windows users, click on '[install R for the first time](https://cran.r-project.org/bin/windows/base/)' and follow the prompts.

**RStudio**

You will also need to download RStudio Desktop (or simply RStudio), which is an integrated development environment to help you write code in R more easily. To download RStudio, follow this [link](https://posit.co/download/rstudio-desktop/) and scroll down to the section titled 'All Installers and Tarballs'. Download the appropriate installer file according to your Operating System. Then, open the installer and follow the installation instructions that you will be prompted with.

**Quarto**

Download Quarto from this [website](https://quarto.org/docs/get-started/). Quarto is a publishing system that will allow you to open and work on the computational notebooks for the workshop. On 'Step 1' on the website, download the version of Quarto that matches your Operating System. Open the installer file, run it and follow the prompts.

**R packages**

Once you have installed R, you will need to install some R extensions, known as packages, that will be useful for the applications explored in this workshop. The packages you need to install are:

-   `tidyverse`
-   `ggthemes`
-   `zoo`
-   `mice`
-   `patchwork`
-   `viridis`
-   `tmap`
-   `sf`
-   `sp`
-   `stringr`
-   `RColorBrewer`
-   `showtext`
-   `scales`
-   `leaflet`
-   `tmap`
-   `mapdeck`
-   `plotly`
-   `htmlwidgets`
-   `nomisr`
-   `flexdashboard`
-   `ggfortify`
-   `forecast`
-   `tseries`

To install the packages, open RStudio. On the console window (normally at the bottom left), write the following command: `install.packages("name of package")`. Make sure you replace "name of package" by the actual name of the package that you want to install e.g. `install.packages("tidyverse")`. Then, press enter and repeat this process until you have installed all the packages in the list.

You can install all the packages by copying and running the code below:

```{r}
#| eval: false
list.of.packages.cran <- c(
   "tidyverse",
   "ggthemes",
   "zoo",
   "mice",
   "patchwork",
   "viridis",
   "tmap",
   "sf",
   "sp",
   "stringr",
   "RColorBrewer",
   "showtext",
   "scales",
   "leaflet",
   "tmap",
   "mapdeck",
   "plotly",
   "htmlwidgets",
   "nomisr",
   "flexdashboard",
   "ggfortify",
   "forecast",
   "tseries"
)

new.packages.cran <- list.of.packages.cran[!(list.of.packages.cran %in% installed.packages()[,"Package"])]
if(length(new.packages.cran)) install.packages(new.packages.cran)

for(i in 1:length(list.of.packages.cran)) {
  library(list.of.packages.cran[i], character.only = T)
}
```

You can load all the packages by copying and running the code below:

```{r}
#| eval: false
deps <- list(
   "tidyverse",
   "ggthemes",
   "zoo",
   "mice",
   "patchwork",
   "viridis",
   "tmap",
   "sf",
   "sp",
   "stringr",
   "RColorBrewer",
   "showtext",
   "scales",
   "leaflet",
   "tmap",
   "mapdeck",
   "plotly",
   "htmlwidgets",
   "nomisr",
   "flexdashboard",
    "ggfortify",
   "forecast",
   "tseries"
)

for(lib in deps){library(lib, character.only = TRUE)}
```

::: callout-note
We might ask you to install more packages on the day that this workshop is taking place.
:::

::: callout-important
Instructions by email will be sent before the workshop to download the Github repository and data which will be used during the workshop. Please ensure you have downloaded the repository and data *before* the workshop.
:::

## During the workshop

All the workshop material will be made available on this [website](https://fcorowe.github.io/lcr-training/) which is currently under construction. Further instructions on how to download the material will be given during the workshop.
